Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-03-28 src/tex/ms.tex
[showyourwork] overleaf sync

Surface liquid water is closely related to planetary habitability: water is required for life as we know it, and life on exoplanets will most likely on be detectable if it is present at the surface. 
To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which an Earth-like planet could maintain surface liquid water \citep{kasting93}.  The HZ is convenient because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. We would therefore like to establish the presence of liquid water on the surface of an exoplanet before trying to interpret its atmospheric biosignatures. %Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., Titan's methane lakes;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.
Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which produce waves and hence increase the size of the glint spot \citep{2011Icar..211..722B,kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean, proving that the method works in practice for an unresolved world.
%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

@@

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.
\cite{Zugger_2010} simulated photometric and polarized phase variations for Earth-like planets including the effects of partial cloud coverage in suppressing the Rayleigh scattering and ocean glint polarization. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. To suppress the effects of Rayleigh scattering, glint observations can be made at longer wavelengths
\citep{Zugger_2011}.

@@

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ is this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:

\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}

\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The on-sky angular separation as viewed by the observer is $\delta$. The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

\subsection{Simulating contrast curves}
\textbf{Kim and David}
%
A second parameter that is crucial for detecting features like rainbows and ocean glint on Earth-like planets is the contrast, i.e. the relative flux of the reflected light from the planet divided by the stellar flux. 
%
While this is not the main focus of this paper, it is important to consider that the contrast and polarised contrast are not constant as function of scattering phase angles.
%
This is due to a combination of the changing illumination fraction of the exoplanets Earth facing surface and scattering effects.
%
Here, we aim to present some simplified examples of the impact of the scattering angle dependent contrast on the detectability of rainbow and glint features. 
%
We note that these calculations are by no means complete or exact, given the many variables that go into modelling reflected fluxes and polarization fractions of an Earth-like planet \citep{ treesstam2019,trees2022}.

>>>>

Surface liquid water is closely related to planetary habitability: water is required for life as we know it, and life on exoplanets will most likely on be detectable if water is present at the surface. 
To first order, the presence of surface liquid water can be predicted by the flux a planet receives from its star. The habitable zone (HZ) describes the range of semi-major axes around a given main sequence star for which an Earth-like planet could maintain surface liquid water \citep{kasting93}.  The HZ is convenient because a planet can be deemed habitable based on solely on its orbital separation, but it remains an untested hypothesis. We would therefore like to establish the presence of liquid water on the surface of an exoplanet before trying to interpret its atmospheric biosignatures. %Ideally, we will empirically establish the presence of liquid water on an exoplanet before trying to interpret its putative biosignatures.

@@

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved reflected light photometry to identify water oceans on the surface of an exoplanet.  \cite{2008Icar..195..927W} performed simulations of scattered light from directly imaged terrestrial exoplanets. 
%to demonstrate that phase-resolved photometry and polarization of a directly-imaged exoplanet would betray the presence of large surface oceans. 
They showed that oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases. %Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but
Specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., Titan's methane lakes;][]{2010GeoRL..37.7104S}. %\citet{lustig2019} explored mapping exoplanets at crescent phases using ocean glint.
Specular reflection appears at similar phase angles in both unpolarized and polarized light and its peak orbital phase can be extended by atmospheric winds, which produce waves and hence increase the size of the glint spot \citep{2011Icar..211..722B,kopparla2018}.
%\cite{2006astro.ph.10518M} quantified how photometric and polarimetric scattered light orbital phase variations might betray the presence of oceans on terrestrial exoplanets.  
\citet{Takahashi2021} used polarimetric Earthshine measurements to detect the presence of the ocean, proving that the method works in practice for an unresolved world.

\noindent {\color{Green}<Modifications by Victor Trees>}

%\cite{2008Icar..195..927W} simulated photometric and polarimetric phase curves of directly imaged terrestrial planets. They noted that surface oceans would be detectable because they are more reflective and exhibit a peak in polarization at crescent phases.

@@

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint {\color{Green} in the total (polarized plus unpolarized) reflected flux}. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.

\cite{Zugger_2010} simulated photometric and polarized phase variations {\color{Green} in the visible wavelength range} for Earth-like planets including the effects of %partial cloud coverage
{\color{Green} Rayleigh scattering and a horizontally homogeneous cloud cover. They concluded that detection of an ocean glint signature in polarization is severely limited by the presence of clouds.} %Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they %again 
%find that oceans are detectable and distinguishable from other sources of polarization. 
{\color{Green} \citet{treesstam2019} simulated the total and polarized fluxes and the degree of polarization of light reflected by ocean planets with a horizontally inhomogeneous (i.e. patchy) cloud cover. They found that an alternation of a cloud-free and cloud-covered glint, as the planet rotates and its cloud cover changes in time, can result in strong fluctuations in the degree of polarization phase curves.}

{\color{Green}In the near-infrared, the Rayleigh scattering optical thickness of the exoplanet's atmosphere is smaller than in the visible wavelength range, which improves the visibility of the glint \citep{Zugger_2011}. Combining degree of polarization phase curves of ocean planets at short wavelengths and long wavelengths results in an intersection of those phase curves, as shown by \citet{treesstam2019}. This color change in polarized light, when the planet orbits its parent star, is not expected for cloudy and cloud-free dry planets, and can therefore help to distinguish between an exo-ocean and forward scattering clouds.}

%Moreover, Trees & Stam (2019) combined the phase curves at various wavelengths between 350 and 865 nm, to show that the glint increasingly reddens the degree of polariza- tion and polarized flux with increasing phase angle. This color change in polarized light does not happen for dry planets and can thus identify an exo-ocean, also in the mean signal of heav- ily cloudy ocean planets with random patchy cloud patterns, and when surface pressures are higher than on Earth.

%The spectra of Stam (2008) and the phase curves of Zugger et al. (2011, 2010) are for horizontally homogeneous planets: the local atmosphere-surface systems do not vary in the latitu- dinal and longitudinal directions. The spectra of partly cloudy planets were mimicked by weighted sums of completely cloudy and completely cloud-free planets, which results in fading of the ocean signatures. Zugger et al. (2010, 2011) therefore concluded that the detectability of an exo-ocean is strongly limited by the presence of clouds.

%Trees & Stam (2019), however, simulated the total and polarized fluxes and the degree of polarization of light reflected by planets with rough ocean surfaces and with horizon- tally inhomogeneous clouds, and found that the glint appearing and hiding behind patchy clouds results in a variability of the phase curves.

%Moreover, Trees & Stam (2019) combined the phase curves at various wavelengths between 350 and 865 nm, to show that the glint increasingly reddens the degree of polariza- tion and polarized flux with increasing phase angle. This color change in polarized light does not happen for dry planets and can thus identify an exo-ocean, also in the mean signal of heav- ily cloudy ocean planets with random patchy cloud patterns, and when surface pressures are higher than on Earth.

@@

{\color{Green} Victor Trees: Do you mean:
%
\begin{equation}
    \Delta \varphi 
    = \varphi_\mathrm{max} - \varphi_\mathrm{min}
    =  2(\varphi_\mathrm{max} - 90^\circ) 
    =  2(90^\circ-\varphi_\mathrm{min} ) \notag
\end{equation}
%
}

The on-sky projected planet distance, $r_\mathrm{proj}$ (in \si{\au}) is shown in \cref{fig:scattering-angle} assuming a non-eccentric circular orbit with a semi-major $a$ (in \si{\au}) at a distance $d_*$ (in \si{\parsec}) from the observer. 
%
The scattering phase angle $\varphi$ in this set up follows the equation,
\begin{equation}
    \sin(\varphi) = \frac{r_\mathrm{proj}}{a}
\end{equation}
%
Using small angles, $r_\mathrm{proj} = \delta d_*$ where $\delta$ is the angular separation on-sky.
%
We access the minimum scattering angle when the projected distance reaches the inner working angle separation $\delta = \mathrm{IWA}$. 
%
Using \cref{eq:Delta_phi}, we deduce:
%
\begin{equation}
    \label{eq:scattering_angle}
    \cos\left(\dfrac{\Delta \varphi}{2}\right) = \frac{\mathrm{IWA \cdot d_*}}{a}
\end{equation}
 %
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The on-sky projected planet distance, $r_\mathrm{proj}$, of a planet with a semi-major $a$ at a distance $d_*$. The on-sky angular separation as viewed by the observer is $\delta$. The scattering phase angle, $\varphi$, is the angle between the line of sight to the star and the line connecting the star and planet.  
    }
    \label{fig:scattering-angle}
\end{figure}

@@

\subsection{Simulating contrast curves}
\textbf{Kim and David}
%
A second parameter that is crucial for detecting features like rainbows and ocean glint on Earth-like planets is the contrast, i.e. the relative flux of the reflected light from the planet divided by the stellar flux. 
%
While this is not the main focus of this paper, it is important to consider that the contrast and polarised contrast are not constant as function of scattering phase angles.https://www.overleaf.com/project/63f87efaa38311f6a4b2a551
%
This is due to a combination of the changing illumination fraction of the exoplanets Earth facing surface and scattering effects.
%
Here, we aim to present some simplified examples of the impact of the scattering angle dependent contrast on the detectability of rainbow and glint features. 
%
We note that these calculations are by no means complete or exact, given the many variables that go into modelling reflected fluxes and polarization fractions of an Earth-like planet \citep{ treesstam2019,trees2022}.
